---
title: "RR v HR Rmd"
author: "Chase Latour"
date: "3/9/2023"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Call needed libraries
library(tidyverse)
library(survival)
library(haven)
library(muhaz)
library(survival)
library(biostat3)
library(survRM2)
library(dplyr)
library(kableExtra)
```

# Call in the data

Call in the data from the server. These data are publicly accessible through Project Data Sphere but require a signed data use agreement, so we do not provide them here.

```{r}

raw_data <- read_sas("/local/projects/DETECTe/data/raw/mCRC_WT/analysis/cohort.sas7bdat")

```

# Clean the dataset

Here we first check the maximum amount of follow-up in each treatment arm.

```{r}

# Look at the maximum amount of follow-up in each treatment arm.
raw_data %>% 
  group_by(trt) %>% 
  summarize(maximum = max(days)) %>% 
  kbl() %>% 
  kable_styling()

```

The maximum amount of follow-up in the untreated is 1,310 days, while the maximum follow-up in the treated is 1,252. We want the same amount of follow-up, so we limit the follow-up for 1,252 days for each person in the analysis.

```{r}

# Want to have a uniform amount of follow-up across the treatment groups, particularly for the non-parametric analyses.
mCRC <- raw_data %>% 
  mutate(days = ifelse(days > 1252, 1252, days),
         event = ifelse(days > 1252, 0, event))

```


# Calculate final HR

We use a Cox proportional hazards model to calculate the HR estimate at 1,252 days of follow-up.

**HR = 0.82 (95% CI: 0.71, 0.93)**

```{r}

cox <- coxph(formula = Surv(days, event) ~ trt, data=mCRC)

summary(cox)



```

# Calculate final RR

We want to calculate the RR at 1,252 days of follow-up for comparison with the HR.

## Function to calculate the RR at a specific point in FUP

This function allows users to calculate the risk ratio at user-specified point in follow-up: `fup_days`.

```{r}

RR_calc <- function(fup_days, km_model){
  
  summary_km <- summary(km_model, times = fup_days)
  
  kmdat <- data.frame(summary_km[c("surv","time","std.err","strata")])
  
  data <- kmdat %>% 
    mutate(r = 1-surv,
           RR = r/lag(r), 
           r0= first(r), r1 = last(r),
           se0 = first(std.err), se1 = last(std.err),
           selnrr = sqrt((1/r0)^2 * se0^2 + (1/r1)^2*se1^2),
           LCL = exp(log(RR) - 1.96*selnrr),
           UCL = exp(log(RR) + 1.96*selnrr)) %>%  
    filter(RR == last(RR)) %>% 
    dplyr::select(c(RR, LCL, UCL)) #%>% 
    #kbl() %>% 
    #kable_styling()
  
  return(data)
  
}

```

## Calculate RR at 1,252 Days

We calculate the RR at 1,252 days here.

```{r}

km <- survfit(Surv(days, event) ~ trt, data=mCRC)

RR_calc(1252, km) %>% 
  kbl() %>% 
  kable_styling()

```

## Calculate RR at different points in follow-up.

We want to plot RR estimates at different points in follow-up. We calculate them here but do not output them. They will be displayed on the survival curve plot.

```{r}

rr_90 <- RR_calc(90, km) # Calculate the RR at 90 days
rr90 <- as.double(round(rr_90[1,1],2)) # Extract the RR

rr_630 <- RR_calc(630, km)
rr630 <- as.double(round(rr_630[1,1],2))

rr_990 <- RR_calc(990, km)
rr990 <- as.double(round(rr_990[1,1],2))

```

# Survival Curve

We plot the study's full survival curve.

```{r}

plot(km, 
     fun = function(x) 1-x,
     lty = c("solid", "dashed"), 
     col = c("#56B4E9","#D55E00"), 
     xlab = "Time in Days From the Beginning of Follow-up",
     ylab = "Cumulative Risk",
     main = "Cumulative Incidence Curve via Kaplan-Meier Estimator",
     cex = 1.3,
     cex.axis = 1.3,
     cex.lab = 1.3,
     cex.main = 1.3,
     lwd = 2,
     mark.time = T)
abline(v=90, lty="dashed")
abline(v=630, lty="dashed")
abline(v=990, lty="dashed")
text(x=200, y=0.08, paste("RR = ", rr90))
text(x=740, y=0.7, paste("RR = ", rr630))
text(x=1100, y=0.85, paste("RR = ", rr990))
legend("bottomright", c("Chemo Only","Panitumumab+Chemo"), 
       lty = c("solid","dashed"),
       col = c("#56B4E9","#D55E00"),
       cex = 1,
       lwd=2)

```




# Plot Kernel-Smoothed Hazard Curves

## Function to Calculate HR at Different FUP

To calculate the HR at different points in follow-up, it is necessary to reset the point at which individuals are administratively censored. This function administratively censors everyone at a user-specified amount of time: `fup_days`.


```{r}

HR_FUP <- function(data, fup_days){
  
  data2 <- raw_data %>% 
    mutate(event = ifelse(days > fup_days, 0, event),
           days = ifelse(days > fup_days, fup_days, days))
  
  cox2 <- coxph(formula = Surv(days,event) ~ trt, data=data2, ties = "efron")
  
  
  return(as.double(summary(cox2)$coefficients[1,2]))
  
}


```


## Calculate HRs at Different FUPs

We calculate the hazard ratios at 90, 630, and 990 days of follow-up. These will not be output here but instead plotting on the kernel-smoothed hazard curves.

```{r}

hr90 <- round(HR_FUP(raw_data,90),2)

hr630 <- round(HR_FUP(raw_data,630),2)

hr990 <- round(HR_FUP(raw_data,990),2)

```

## Create Plot

Calculate the kernel-smoothed objects.

```{r}

trt0.mCRC <- muhaz(mCRC$days, mCRC$event,subset = mCRC$trt == 0)
trt1.mCRC <- muhaz(mCRC$days, mCRC$event,subset = mCRC$trt == 1)

```

Calculate the approximately the HRs at each time point via the non-parametric estimates.

```{r}

#Determine which location on the vector of time points is closest to the estimates of interest

## TRT = Chemo Only
est90_0 <- max(which(trt0.mCRC$est.grid <= 90))
est630_0 <- max(which(trt0.mCRC$est.grid <= 630))
est990_0 <- max(which(trt0.mCRC$est.grid <= 990))
## TRT = Panitumumab+Chemo
est90_1 <- max(which(trt1.mCRC$est.grid <= 90))
est630_1 <- max(which(trt1.mCRC$est.grid <= 630))
est990_1 <- max(which(trt1.mCRC$est.grid <= 990))

# Get the Hazards at that point

## Trt = Chemo Only
h90_0 <- trt0.mCRC$haz.est[est90_0]
h630_0 <- trt0.mCRC$haz.est[est630_0]
h990_0 <- trt0.mCRC$haz.est[est990_0]
## Trt = Chemo + Panitumumab
h90_1 <- trt1.mCRC$haz.est[est90_1]
h630_1 <- trt1.mCRC$haz.est[est630_1]
h990_1 <- trt1.mCRC$haz.est[est990_1]

# Finally, the kernel-smothed HRs
h90 <- h90_1/h90_0
h630 <- h630_1/h630_0
h990 <- h990_1/h990_0

```

This provides a non-parametric estimate of the hazard rate curves in the two groups.

```{r}

plot(trt0.mCRC, xlab="Time in Days From the Beggining of Follow-up")
lines(trt1.mCRC, lty="dashed")
segments(x0= 0,x1=90, y0= 0.000, y1=0.000, lwd=1.5, lty="dashed", col="red")
segments(x0=0, x1=630, y0=0.00022, y1 = 0.00022, lwd=1.5, lty="dashed", col="red")
segments(x0=0, x1=990, y0=0.00055, y1=0.00055, lwd=1.5, lty="dashed", col="red")
legend("bottomright", c("Chemo Only","Panitumumab+Chemo"), 
       lty = c("solid","dashed"),
       cex = 0.8,
       lwd=2)
text(x=225, y=0.000, paste("PH HR = ", hr90))
text(x=500, y= 0.00035, paste("PH HR = ",hr630))
text(x=900, y= 0.0007, paste("PH HR = ", hr990))
segments(x0=90, x1=90, y0=0.0011, y1=0.004, lty="dotted")
segments(x0=630, x1=630, y0=0.0011, y1=0.004, lty="dotted")
segments(x0=990, x1=990, y0=0.0011, y1=0.004, lty="dotted")
text(x = 205, y=0.0015, paste("KS HR = ", round(h90,2)))
text(x = 745, y=0.0035, paste("KS HR = ", round(h630,2)))
text(x = 1052, y=0.0021, paste("KS HR = ", round(h990,2)))



```



# Interaction Term in Cox Model

Model with interaction term at 700 days.

```{r include=FALSE}

interaction <- coxph(Surv(days,event) ~ trt + tt(trt), data = mCRC, ties = "efron", tt = list(function(trt, days, ...){trt*days>700}))

summary_int <- summary(interaction)

CI_est <- summary_int$conf.int

coefficients <- summary_int$coefficients

logHR <- as.double(coefficients[1,1]) + as.double(coefficients[2,1])
varHR <- as.double(coefficients[1,3])**2 + as.double(coefficients[2,3])**2
q <- qnorm(0.975)
LCL <- exp(logHR - q*sqrt(varHR))
UCL <- exp(logHR + q *sqrt(varHR))
HR <- exp(logHR)


# HR from 0- <= 700 days: 0.79, SE=0.07063, coef=-0.23797

l700 <- paste(round(CI_est[1,1],2), " (", round(CI_est[1,3],2),", ", round(CI_est[1,4],2),")")

# HR from 700 on: 

g700 <- paste(round(HR,2)," (",round(LCL,2),", ",round(UCL,2),")")

data.frame(Timing = c("Less than 700 days", "From 700 days"),
                                HR = c(l700, g700)) %>% 
  kbl() %>% 
  kable_styling()

```


Model with interaction term between treatment and continuous time.

```{r}

continuous <- coxph(Surv(days,event) ~ trt, 
                    data = mCRC, 
                    ties = "efron")
continuous2 <- cox.zph(continuous, transform = "identity")



# Plot the time-varying coefficient for treatment

plot(continuous2[1],
     ylim=c(-1.1, 1.1),
     xlab = "Time in Days from Start of Follow-up",
     ylab = "Natural Log Transformed Hazard Ratio"
     )
abline(0,0, col="grey", lty=3, lwd=1)
abline(h=continuous$coefficients[1], col=3, lwd=2, lty=2)
legend("bottomright",
       legend=c("Traditional Cox HR",
                "Time-varying HR"),
       lty=c(2,1), col=c(3,1), lwd=c(2,2))



```


## Restricted Mean Survival Time

Here, we find the following:

**Trt 1:** 344.6 (318.4, 370.8) Days

**Trt 0:** 289.4 (264.8, 314.0) Days

**Difference:** 55.2 (19.2, 91.1) Days

```{r echo=TRUE}

rmst2(mCRC$days, mCRC$event, mCRC$trt, tau = 1095)

## Trt 1: 344.6 (318.4, 370.8)
## TRT 0: 289.4 (264.8, 314.0)

## Difference: 55.2 (19.2, 91.1)

```



